Optimizing the future of biodiversity sampling by citizen scientists

### 2019-04-10 paper/main.tex
More more more!

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{The authors have no conflict of interest to declare.}

@@

Citizen science projects need to continuously adapt to be successful, and to better inform conservation science. Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.

The same approach is robust to spatial scale: whether a citizen science project is global, regional, or carried out in the constraints of a local park, the site can be gridded and observations can then be guided into the future.

We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day. For every day, a citizen scientists can be presented with a map, showing the highest valued sites that should be visited. Imagine visitor centers at National Parks and

Feedback to citizen science participants is important! Imagine visitor centers across the world at national parks or urban greenspaces providing users with a localized map showing which trail someone should visit if they are interested in contributing to citizen science. More broadly, this approach could be applied to anyone in the world. The global pull of ecotourism is increasing exponentially, allowing the the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled.

This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations.

Here, our framework focused on a specific statistical outcome: trend detection. Many other ecological outcomes arise from citizen science, including species distribution models, phylogeographical research, invasive species detection, or phenological research. These potential outcomes will have different optimal sampling strategies in space and time. We note that our framework is robust, with the necessary piece of information being a statistical model to extract leverage.

This approach relies on statistical leverage, assigning every citizen science observation a measure of influence that influences a statistical model.

- Looking forward but using the past
- We only looked at trend detection, but many other models can be used in this process (e.g., species distribution models)
- Other potentials for this general approach (identifying outliers in the dataset or misidentified species)

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

\subsection*{Study region} We illustrate our concepts throughout the Greater Sydney Region, and we gridded the region into 411 grids, all of which are ~ km. We used the R statistical environment to carry out all analyses.

\subsection*{eBird data} We downloaded the eBird basic dataset (VERSION), and subsetted the data between January 1st, 2010 to December 31st, 2018. eBird collects data in the form of 'checklists' --- a list of all species identified (audibly or visually) for a given spatiotemporal coordinates. eBird relies on an extensive network of regional reviewers who are local experts of the avifauna [REF]

\subsection*{Trend detection model} For illustration purposes, we fitted a simple generalized linear model, based on presence/absence for each of 274 species --- those of which had > 20 observations in the region throughout the study period. The GLM took the form of: XXXXXX. County was included to add a spatial representation throughout the Greater Sydney Region, while we included the length of the list in the model as an offset to account for the effort associated with each individual eBird checklist.

We acknowledge that a number of different models can be fit which assesses trend detection [REFS], and biases can be accounted for in a variety of ways [REFS]. These models were fitted for the period between 2010-2018.

\subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable [REFS]. In other words, it is a measure of how much an observation influences the results of an observation. In our instance, because we had multiple predictor variables in our GLMs, we used dfBeta as a measure of statistical leverage. dfBeta can be calculated for a given parameter in a regression model.

For each species (N=274) we calculated the dfBeta for each of the 29,055 observations used for each model. Then for every checklist assessed in the model, we took the sum of the absolute value of the dfBetas for that checklist as a measure of a checklist's leverage in understanding species' trends throughout the Greater Sydney Region.

\subsection*{Parameter calculation} For each day in 2018 (N=365), for each grid, we dynamically calculated the following parameters, based on sampling: (1) whether a grid cell was sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date. These are some hypothesized parameters which we felt would influence the relative value of a given sampling event.

We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids. We ran a linear regression to parameterize what parameters had the greatest impact on the leverage of a given observation.

\subsection*{Expected leverage calculation} We XXXXX.

\subsection*{Data availability}

}

>>>>

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: C.T.C., W.K.C. designed research, performed research, and analyzed research. C.T.C., W.K.C., J.J.L.R., A.G.B.P., and R.E.M. wrote the paper.}
\authordeclaration{The authors declare no conflict of interest.}

@@

We found generally consistent results, based on the grid size chosen, highlighting the influence of spatial scale of our approach. Ultimately, this approach could be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park; the site can be gridded and observations can then be guided into the future. It is likely that the spatial grid cells would need to change, relative to the spatial scale of the study.

Citizen science projects need to continuously adapt to be successful, and to better inform conservation science. Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions. We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day. Feedback to citizen science participants is important! This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations. Imagine visitor centers across the world at national parks or urban greenspaces providing users with a localized map showing which trail someone should visit if they are interested in contributing to citizen science. More broadly, this approach could be applied to anyone in the world, and citizen science participants of a given project could opt in to receive daily, weekly, or monthly updates on where they should sample. The global pull of ecotourism is increasing exponentially, allowing the the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled.

Here, our framework focused on a specific statistical outcome: trend detection. Many other ecological outcomes arise from citizen science, including species distribution models, phylogeographical research, invasive species detection, or phenological research. These potential outcomes will have different optimal sampling strategies in space and time. For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids than for species trend detection. We note that our framework is robust, with the necessary piece of information being a statistical model to extract leverage.

Arguably, citizen science will continue to shape the future of ecology and conservation, as it has substantially for the past 100 years [REFS]. But we need to look forward the future: how can we maximize the immense volunteer effort? In eBird alone, citizen scientists contributed XXXX number of volunteer hours focused on looking for birds. Ultimately, birders do a good job of sampling, but we believe that we can still optimize the collective citizen science effort. We provide a framework with which citizen science project managers can use fit-to-purpose.

\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km^{2} (Fig. SX), where a grid represented a 'site'. We used the R statistical environment \cite{rcoreteam2018r} to carry out all analyses, relying heavily on the tidyverse \cite{wickham2017tidyverse}, ggplot2 \cite{wickham2016ggplot}, and sf \cite{pebesma2018sf} packages.

In order to test our predictions, we relied on the eBird basic dataset (version ebd_relDec-2018; available at: <https://ebird.org/data/download>), subsetting the data between January 1^{st}, 2010 to December 31^{st}, 2018. eBird is a particular successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}. eBird relies on volunteer birdwatchers who collect data in the form of 'checklists' --- a list of all species identified (audibly or visually) for given spatiotemporal coordinates. eBird relies on an extensive network of regional reviewers who are local experts of the avifauna \cite{gilfedder2019brokering} to ensure data quality \cite{sullivan2009ebird}.

\subsection*{Trend detection model} We first filtered the eBird basic dataset \cite{callaghan2017assessing, johnston2018estimates, la2014role}, by the following criteria: (1) we only included complete checklists, (2) only included terrestrial bird species, (3) removed any nocturnal checklists, (4) only included checklists which were > 5 minutes and < 240 minutes in duration, (5) only included checklists which travelled < 5 km or covered < 500 Ha, and (6) only included checklists which had > 4 species on it, as checklists with less than 4 species were likely to be targeted searches for particular species \cite{walker2017using, szabo2010regional}.

For any species with > 50 observations (N=235), we fit a generalized linear model using the 'glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}. The models consisted of a continuous term for day, beginning on January 1^{st}, 2010, and a categorical term for county, providing a spatial component to the models. We also included an offset term for the number of species seen on a given eBird checklist, accounting for temporal and spatial effort of that checklist \cite{szaboe2010regional}. Each of the models were fit with a binomial family distribution. A total of 25,995 observations (i.e., eBird checklists) were used to fit each of the models.

\subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable \cite{cook1977detection}. In other words, it is a measure of how much a given observation influences the results of a statistical outcome. In our instance, we had multiple predictor variables in our GLMs, and so we used dfBeta as a measure of statistical leverage for each observation. dfBeta measures the change to the observed parameters of a new regression equation, after omitting the \textit{i^{th}} observation from the dataset \cite{belsley1980regression}. It follows the formula:

\begin{equation}
D F B E T A=\hat{\beta}-\beta_{(i)}=\frac{\left(X^{\prime} X\right)^{-1} X_{i} r_{i}}{1-h_{i}}
\end{equation}

where X is th predictor variable matrix, \textit{r} the residual vector, \textit{i h} the \textit{i^{th}} diagonal member, and \textit{i x} the \textit{i^{th}} line of matrix X. The value of dfBeta proportionally decreases with increasing number of observations.

In our case, each observation for a given species recieved a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the 'dfbetas' function from R \cite{rcoreteam2018r}. The measure of statistical leverage, then, of a given checklist was the sum of the absolute value of the dfBeta measures for each species (i.e., the sum of all 235 dfBetas). This measure of statistical leverage was thus a measure of a checklist's influence in understanding cumulative species' trends throughout the Greater Sydney Region.

\subsection*{Parameter calculation} After our model was fit from 2010---2018, we calculated the predicted parameters of interest for each day in 2018 (N=365). For each individual grid, at each of the grid sizes, we dynamically calculated the following parameters: (1) whether a grid cell had ever been sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date, and (7) the number of unique sampling days within the grid cell.

We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids on that day. We ran a linear regression, for each of the different grid sizes considered in the analysis, to investigate what parameters were of significant interest. Prior to modelling, duration was highly correlated with median sampling interval for the majority of the grid size analyses, and as such, was excluded from consdieration. Given the paramters' correlation varies among grid cell sizes, we need to ensure a robust, and simple model. All variables were standardized prior to modelling, ensuring that the effect sizes of the given parameters were meaningful. The response variable, dfBeta (i.e., value) was log-transformed prior to modelling to meet model assumptions. Thus, the final model included a log-tranformed dfBeta response variable, regressed against standardized median sampling interval, number of days sampled, days since the last sample, distance to the nearest sampled neighbor, and the neighbor's median sampling interval.

After our model was fit, we used the 'augment' function from the broom package \cite{robinson2018broom} to predict the expected leverage for every grid cell in the Greater Sydney Region, for every day. For grid cells which were unsampled, we assigned them the mean of the sampled grid cells, based on our lack of evidence that unsampled cells were significantly more valuable than sampled cells. This prediction was done for every day of 2018.

\subsection*{Data availability} eBird data are available for download, but the necessary portion of the dataset, along with spatial data, and code to reproduce our analyses are available at: GitHub repository.
}
